---
title: "Updated Method for Analyzing Survival in Entangled Right Whales"
author: "Rob Schick"
date: "February 11, 2018"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(dplyr)
library(tangled)
```

The purpose of this document is to provide an initial overview of the suggested changes for analyzing the survival output from the right whale model. Our goal is to summarize survival by injury -- in order of increasing severity:

1. Severe with gear
2. Moderate with gear
3. Severe without gear
4. Minor with gear
5. Moderate without gear
6. Minor without gear

...and ideally to do this for both males and females. For now, we won't investigate the effects of age..

Though we had previously done this using the the Kaplan-Meier apprach, Jim now feels that is the incorrect metric because you have to know who survived. I had previously implemented this using the right censoring approach, but he doesn't want to go that route. Instead, what he proposes is to build up comparisons between the posterior estimate of survival probability for survival of the entangled animal - following the end of the last event, and compare it to the mean survival probability of reference animals during the period from the end of the entanglement until the survival probability goes to 0.

The algorithm would look something like this.

1. We start by extracting the last entanglement event from this animal
2. From that we gather the time information corresponding to the end of the window
3. Calculate the survival probability from the end of entanglement window until presumed death

We start by assembling and viewing a data frame of all of the last entanglement event for each animal:

```{r lastEvents}
tangEvents <- tangleOut %>% 
  group_by(EGNo) %>% 
  # summarise(nevents = max(EventNo))
  filter(EventNo == max(EventNo)) %>% 
  arrange(EGNo)
```

For the paper, we also excluded animals that were ship struck, or had severe prop scars. This list is below:

```{r}
exclude <- c(1004, 1006, 1014, 1045, 1128, 1223, 1267, 1308, 1504, 1623, 1907, 1909, 2143, 2150, 2220, 2250, 2404, 2425, 2450, 2617, 2820, 3508, 3522, 3853)
tangSub <- tangEvents %>% 
  filter(!(EGNo %in% exclude))
```


## Entangled Animal
We'll start with the first animal to illustrate the algorithm; once complete we'll run it for all animals. 

```{r}
event1 <- tangSub[1, ]
event1
```

This particular animal had one event with a very long possible window. We shortened that to be approximately 4 months: 

* Start of window: ```r event1$StartDateWindow```
* End of window: ```r event1$EndDateWindow```

We need to find its survival probability over that window:

```{r}
idx <- which(ID == event1$EGNo)
# deathyr[idx, ] 
svec <- 1 - cumsum(deathyr[idx, ] / ng) 
svec[1:firstSight[idx]] <- NA
# plot(svec)

```

Now we want to define the window and view the survival probability over that window.

```{r}
w <- which(deathyr[idx, ] > 0)
w1 <- min(w)
w2 <- max(w)
theta <- svec[(w1 - 1):w2]
testVar <- theta * (1 - theta)
plot(theta, type = 'l', ylab = 'Survival Probability', 
     xlab = 'Months Following End of Entanglement',
     main = paste0("Sighting EGNo: ", event1$EGNo))
lines(theta + testVar, lty = 2)
lines(theta - testVar, lty = 2)
```


## Reference Animal
Ok, with that built for the entangled animal, we want to sample the reference animals - against which we'll compare survival. In practice that means all animals who are alive for the whole duration of the window and who are not entangled during that interval. We'll summarize the mean survival of all those animals for each of the months during the entanglement survival window. Let's say we have a an 8-month survival window when the entanglement declines from 1 to 0. During each of those months, we'll calculate the difference in survival between the individual animal and the reference class. Each of these animals will have metadata that indicates injury and gender. By building up the distribution of differences we will be able to quantify the impact of entanglement on survival by category. Note that we will only compare females to females, and males to males; specifically I mean for an entangled female, the reference class will be comprised of all females.

```{r}
load("E:/rob/Documents/research/projects/rightWhaleEntanglement/src/data-raw/eg_203_ng_50000_BIG_25000_BIG_25000.rdata")
ecand <- tangleOut[year(tangleOut$StartDateWindow) != endWinDate, 'EGNo']
ecand <- unique(ecand[gender[match(ecand, ID)] == 'F']) 
ssub <- sights[sights$GenderCode == 'F' & sights$SightingYear == endWinDate, ]
ocand <- unique(ssub$SightingEGNo)
fincand <- ocand[!(ocand %in% ecand)]
fincand
```

The processing logic of the previous code block is as follows:

1. load in a completed run of the PCAD model
2. Find animals that a) are not entangled during the year we are examining; and b) are the same sex (this is the ```ecand``` vector)
3. From the original sightings data, find all females that were seen in the year we are examining (```ssub```). (Note this is restrictive; we could expand this to all females assumed alive.)
4. Pare down the sightings data to just have a vector of individual ```SightingEGNo``` values (```ocand```)
5. Find a final candidate list of animals seen that year who were not entangled (```fincand```)

With that list of possible animals, then we can choose one at random, and assemble its health.

```{r}
cID <- sample(fincand, 1)
cidx <- which(ID == cID)

```


### Distribution of Differences
Once we've sampled from the posterior for each animal, then we have to examine the differences.

```{r}
hdiff <- himpMat - hrefMat
hsurvsum <- rowMeans(hdiff)
hist(hsurvsum, breaks = 30, main = 'Difference Between Impacted and Healthy Animal', xlab = 'Health Units')
```

### Which is the Best Reference Animal
One of the issues the above analysis raises, is how do we know that this is a good animal to compare against? What if we chose a different animal, that for whatever reason led to a skewed result. I'll try to investigate that by comparing the health of this impacted animals against a range of candidates to see what the results look like. Se we can repeat the above calculation, but now do it for all candidate animals, not just this one reference.

```{r}
hdiffList <- vector('list', length(fincand))
ndraw <- 1:1000

for (j in seq_along(fincand)) {
  cID <- fincand[j]
  cidx <- which(ID == cID)

  hmomMean <- healthmean[cidx, startW:endW]
  if(any(is.na(hmomMean))) next
  hmomSD <- healthsd[cidx, startW:endW]
  hrefMat <- matrix(NA, max(ndraw), ncol = length(hmomMean))

  for (i in seq_along(ndraw)){
    hrefMat[i, ] <- rnorm(length(hmomMean), hmomMean, hmomSD)
  }

  hdiffList[[j]] <- data.frame(refAn = cID, meanDiff = rowMeans(himpMat - hrefMat))
}
  
```

We need to create a data frame to make it a bit easier to plot.

```{r}
hdiffList2 = hdiffList[-which(sapply(hdiffList, is.null))]
hdiffdf <- do.call(rbind.data.frame, hdiffList2)
```

With that built up, let's plot it.

```{r}
library(ggplot2)
pall <- ggplot(hdiffdf, aes(meanDiff))+
  geom_density()+
  labs(title = 'All Animals Averaged Together')
pall

p <- ggplot(hdiffdf, aes(meanDiff, group = refAn))+
  geom_density()+
  labs(title = 'All Animals Plotted Separately')
p
```



## Impacts of Entanglement on Survival
Ok, with survival the idea is similar -- we take an impacted animal and look at its survival curve in relationship to a randomly sampled reference individual. There are several different ways to do this. We'll start here by looking at probability of survival for the impacted animal during the entanglement event and compare that to a randomly selected animal. (Here, I'll use a different impacted animal from above because 1014 is an observed dead animal - hence no uncertainty in death.) We then build up the distribution of differences in survival and plot them. Again, as before, the selection of the reference case bears careful consideration. One first set of candidates could be the animals that were never entangled, but there could be year/animal mismatches here once we start sampling more. By this I mean with so few animals that bear no scars, we could end up with year and month combinations where we have an impacted animal, but no reference animal.

Let's look at the posterior, which provides an estimated month the animal died in, and then we can use that to summarize survival.

```{r}
anID <- 1130
idx <- which(ID == anID)
deathyr[idx, 360:375] # this snippet shows the candidate months of death
round(deathyr[idx, 360:375] / ng, 4)
svec <- 1 - cumsum(deathyr[idx, ] / ng) 
svec[1:firstSight[idx]] <- 0
# plot(svec)

test <- round(svec[360:375], 4)
testVar <- test * (1 - test)
plot(test, type = 'l', ylab = 'Survival Probability', xlab = 'Months Following Last Entanglement')
lines(test + testVar, lty = 2)
lines(test - testVar, lty = 2)
```

That represents the survival with uncertainty for one animal between the time of the last entanglement and when survival probability goes to 0. We need to sample from this posterior 1,000 times to build up the distribution of survival probabilities. And then take the difference between that and the survival probabilities of the reference animals. First the entangled whale:

```{r}
probs <- test
survMat <- matrix(NA, nrow = 1000, ncol = length(probs))

for(i in 1:nrow(survMat)) {
  survMat[i, ] <- rbinom(prob = probs, n = length(probs), size = 1)
}

# pvec <- apply(survMat, 2, mean)
```

Next, we find the reference animmals, and as above, build up the survival probabilities. Then we can subtract them to get the distribution of differences. For reference, we'll start with the candidates that we identified above. We need to get the correct date/times selected as well to make sure the comparisons are valid. By this I mean that if our window for the entangled animal is January through July of 2000, then we need to select these times for candidate whales.

```{r candidatePsurv}
canSurv <- matrix(data = NA, nrow = 1000, ncol = fincand)


```

