---
title: "Evaluation of Health and Calving Intervals"
author: "Rob Schick"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluation of Health and Calving Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of this vignette is to describe health during the entanglement windows to understand if entanglement impacts health and in turn, reproduction. To do this we will calculate if health dropped below the previously published value of 67 during a period of availability and plot the duration of this against the interval. Our assumption is that longer calving intervals are related to length of time that a reproductively active female spends below that threshold, i.e. more time below the health threshold, the longer the interval is likely to be. We will experiment with other covariates as well, e.g. the health value during the entanglement, a 0/1 factor variable if below, the length of the time period below the threshold (if below), and the entanglement category.

We begin by preparing the data.

## Data Preparation
We have two different data frames that we need to work with: 1) the individual estimates of health that form the core of the model output, and 2) the list of 48 candidate animals for this test.

A reminder that this is the order of the categories:

gearInj | Definition
-----------|----------
1 | Severe with gear
2 | Moderate with gear
3 | Severe no gear
4 | Minor with gear
5 | Moderate no gear
6 | Minor no gear

We take the `tangRepro` data frame that is created in the `prepAmyEntanglementData.R` script as part of `tangled` package construction. We loop over the events and extract monthly estimates of the health anomaly during the event, as well as a derived variable `lthold` that indicates the number of months (if any) the health estimate was below 67. 

```{r global_options, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

library(tangled)
library(tidyverse)
library(ggplot2)
library(gghighlight)
```


```{r}
tmp <- prepBoxplotHealthData(tangRepro, tangNonRepro, anomFlag = FALSE, thold = 67)
dfLong <- tmp$dfLongRepro %>% 
  mutate(severity = case_when(gearInj %in% c(6, 4) ~ 'minor',
                              gearInj %in% c(5, 2) ~ 'moderate',
                              gearInj %in% c(3, 1) ~ 'severe')) %>% 
  arrange(egno, eventNo)
```

At the end of this looping the new data look like this:

```{r, warning=FALSE, message=FALSE}
head(dfLong)
```

We can quickly summarise that to get a feel for how health changes as a function of gear injury type as well as an estimate of how many events are in each category. This simply shows the median health values during the impacted period, as well as the median number of months below a health value of 67.

```{r }
dfLong %>% 
  dplyr::filter(variable == 'impacted') %>% 
  group_by(gearInj) %>% 
  summarise(anomaly = median(hAnom, na.rm = TRUE),
            n_months = median(lthold)) %>% 
  mutate(severity = case_when(gearInj == 6 ~ 'minor',
                              gearInj == 4 ~ 'minor - gear',
                              gearInj == 5 ~ 'moderate',
                              gearInj == 2 ~ 'moderate - gear',
                              gearInj == 3 ~ 'severe',
                              gearInj == 1 ~ 'severe - gear'))
```

## Pairing with Amy's Case Studies

Amy prepared a set of data that 

```{r}
cases <- read_csv(here::here('data-raw', '2020-07-17_Calving-interval-analysis_Case-Study_all-cases.csv')) %>% 
  select(!starts_with('X')) %>% 
  arrange(EGNo, CalvingYear)
```

I start by joining the two databases

```{r}
df_out <- full_join(dfLong, cases, by = c("egno" = "EGNo_1", "eventNo" = "EventNo")) 
```

Now I need to summarize over the same calving year for each animal

```{r}
df_sum <- df_out %>% 
  group_by(egno, CalvingYear) %>% 
  summarize(lthold = sum(lthold),
            health = mean(hAnom),
            gearInj = min(gearInj),
            interval = unique(CalvingInterval)) # to retain the worse event when there are multiple groupings
```

A quick plot

```{r fig.width=6.5}
ggplot(data = df_sum, aes(lthold, interval))+
  geom_jitter() +
  labs(x = "# of Months Below 67", y = 'Calving Interval (years)')
```

And an even quicker `lm()` to see what we see

```{r}
summary(lm(interval ~ lthold, data = df_sum))
```

