---
title: "Entanglement Analysis"
author: "Rob Schick"
date: "`r Sys.Date()`"
output:
    html_document:
      css: custom.css
      toc: yes
vignette: >
  %\VignetteIndexEntry{Entanglement Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This will produce a vignette of the analyses we will conduct for the entanglement paper. The analyses we will build through are as follows:

- Compare Health of entangled vs. non-entangled animals
- Compare health before, during and after entanglement
- Tabulate incidents of entanglement by age/reproductive class and severity
- Examine health scores of entangled whales in relationship to year-specific health threshold (from Rolland et al)
- Plot health and reproduction as a function of median inter-calf interval and whether an animal is over this median
- Tabulate survival from last detection
- Plot this over time by entanglement severity
- Plot cumulative impacts over time of right whales

Each of these will have its own section, as needed, and the links to these sections are provided on the left. Let's get to work.

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Health of Entangled vs. Non-entangled Animals
The goal of this section is to look at the health of entangled whales versus whales that were not entangled. To do this we need a few definitions. 

```{r, child = 'healthAffectedByEntanglements.Rmd'}

```



