---
title: "Survival of Entangled Right Whales - Excluding Those Struck and Killed by Ships"
author: "Rob Schick, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival of Entangled Right Whales - Excluding Those Struck and Killed by Ships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Survival of Entangled Whales be Sex and Severity - Ship Struck Animals Removed
Here we extend the previous analysis to look at survivorship of entangled whales who did not die as a result of being struck by a ship. We'll do this by backing them out of the data frames used to estimate survivorship. The list of animals to be exlcuded is:

* 1128
* 1504
* 1907 (although also a severe with gear engtl)
* 1223
* 2250
* 1623
* 2220
* 2450
* 1014
* 2150
* 1004
* 1909
* 2143
* 2617
* 1267
* 3508
* 1308

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```


```{r,echo=FALSE}
library(tangled)
library(magrittr)
library(dplyr)
```

To get and store just that last event, we call a function called `makeEvents()`, which will prepare a data frame of the last entanglement event of each animal and add some identifying information about its death status (see next section for details). However, as compared to the regular survival analysis, here we'll back out the animals in the list above:

```{r}
events <- makeEvents()
nrow(events)
avec <- c(1128, 1504, 1907, 1223, 2250, 1623, 2220, 2450, 1014, 2150, 1004, 1909, 2143, 2617, 1267, 3508, 1308)
events <- dplyr::filter(events, !EGNo %in% avec)
nrow(events)
```

